Active Learning for SAT Solver Benchmarking

### 2022-10-09 paper/paper.tex
Remove some resolved JB comments in Sections 3 & 4
No impact on visible output.
Fix grammar and spelling in sections 1 to 4

Our approach satisfies several desirable criteria for benchmarking:
It outputs a \emph{ranking} to compare the new solver against the existing set of solvers.
For this ranking, we do not even need to predict exact solver runtimes, which is trickier.
We optimize the \emph{runtime} that our strategy needs to arrive at its conclusion.
We use problem-instance and runtime \emph{features}.
Moreover, we select instances \emph{non-randomly} and \emph{incrementally}.
In particular, we take runtime information from already done experiments into account when choosing the next.
By doing so, we can control the properties of the benchmarking approach such as its required runtime.
Rather than solely outputting a binary classification, i.e, the new solver is worse than existing solver A, we provide a \emph{scoring} function that shows by which margin a solver is worse and how similar it is to existing solvers.
Our approach is \emph{scalable} in that it ranks a new solver $\hat{a}$ among any number of known solvers $\mathcal{A}$.
In particular, we only subsample the benchmark once instead of comparing pairwise against each other solver~\cite{MatriconAFSH21}.
Since a new solver idea is rarely best on the first try, it is desired to obtain a solver ranking fast.
In this way, a new solver idea can be discarded if it performs poorly across the board or may be further tweaked if it shows promising results at least in some cases.

We evaluate our approach with the SAT Competition~2022 Anniversary Track benchmark~\cite{sat2022}, consisting of 5355~instances and full runtimes of 28~solvers.
We perform cross-validation by treating each solver as new once, and predict these solvers' PAR-2 rank.
On average, our predictions reach about \SI{92}{\%} accuracy with only about \SI{10}{\%} of the runtime that would be needed to evaluate these solvers on the complete set of instances.

@@

Hoos et al. discuss which properties are most desirable when selecting SAT benchmark instances~\cite{HoosKSS13}.
The selection strategy presented is static, i.e., it does not depend on particular solvers to distinguish.
Selection criteria are a instance variety to avoid over-fitting, adapted instance hardness (not too easy but also not too hard), and avoiding duplicate instances.
To filter too similar instances, they use a distance-based approach with the SATzilla features~\cite{XuHHL08,features}.
Their approach \emph{ranks} solvers, is \emph{feature-based}, and \emph{scalable} in the sense that the evaluation time scales linearly with more instances and solvers.
Also, it allows \emph{scoring} and thereby comparing solvers.
However, the approach does not optimize for benchmark \emph{runtime} and selects instances \emph{randomly}, apart from constraints on the instance hardness and feature distance.

@@

Matricon~et~al. present an incremental benchmark selection approach~\cite{MatriconAFSH21}.
Their \emph{per-set efficient algorithm selection problem} (PSEAS) is similar to our \emph{New-Solver Problem} (cf.~Definition~\ref{def:new-solver-problem}).
Given a pair of SAT solvers, they iteratively select a subset of instances until the desired confidence level is reached to decide which of the two solvers is better.
The selection of instances is dependent on the choice of the solvers to distinguish.
They calculate a scoring metric for all unselected instances, run the experiment with the highest score, and update the confidence.
Most of the proposed strategies in~\cite{MatriconAFSH21} are instance-based, i.e., no model of the solver runtimes is built.
Regarding our desired criteria of benchmark selection (Table~\ref{tab:requirements}), the aforementioned approach \emph{ranks} solvers, optimizes for \emph{runtime}, is \emph{feature-based}, and uses \emph{incremental non-random} sampling.
However, the approach only compares solvers binarily rather than providing a \emph{scoring}.
Thus, it is not clear how similar two given solvers are or on which instances they behave similarly.
Moreover, a major shortcoming lies is the lacking \emph{scalability} with the number solvers.
Comparing only pairs of solvers, evaluating a new solver requires sampling several benchmarks.
In contrast, our approach allows comparing a new solver against a set of existing solvers with one benchmark.

@@

The posed \emph{New-Solver Problem} has stark similarities to the well-studied field of active learning AL within recommender systems, especially the \emph{new-user problem}~\cite{RubensESK15}.
On the one hand, we want to maximize the utility an instance provides to our model and, on the other hand, minimize the cost (CPU time) that is associated with its acquisition.
In contrast to traditional passive machine-learning methods with given instance labels, active learning allows selecting instances for which to acquire labels.
AL algorithms can be categorized into \emph{synthesis-bas\-ed} \cite{0001AEMN22,GarzonMG22,2019gaal} and \emph{pool-bas\-ed} approaches \cite{GolbandiKL11,HarpaleY08,KorenBV09}.%SinhaED19,distribAL,TongK01,KapoorGUD07,KornerW06,MelvilleM04}.
While synthesis-based methods generate instances for labeling, pool-based methods rely on a fixed set of unlabeled instances from which to sample.

@@

Preliminary experiments, as well as research in portfolio solvers, have shown however that directly predicting runtimes without any transformation does not work well in practice~\cite{NgokoCT19,CollauttiMMO13}.
%JB: Collauti at el. still predict actual (though normalized) runtimes rather than discrete labels, they only heavily discretize runtimes when using them as features
%TF: added "without any transformation"
In our experiments, the mean squared error was within the same magnitude as the values to be predicted.
In contrast to that, discretizing runtimes and using those labels in the model's training and prediction yields good results (cf.~Section~\ref{sec:eval}).
%The following paragraphs motivate why it makes sense to discretize runtimes.%MI: no further motivation found, commented out
%JB: we should probably cut this a bit
%TF: I've commented out a few sentences but we probably need to cut more.

We use a function $\gamma\!: {\mathcal{A} \times \mathcal{I}} \rightarrow \left\lbrace 1, \dots, k \right\rbrace$ to transform runtime observations.
In our experiments, this function is a clustering algorithm.
Note that this transformation is done per instance to adapt to different instance difficulties. 
Also, we define ${\gamma\!\left(a, i\right) := k}$ iff $r\!\left(a, i\right) = \,\perp$ for all $a \in \mathcal{A}$ and $i \in \mathcal{I}$.
To determine a ranking based on those labels, we define
\begin{equation}
  \operatorname{score}\!\left(a\right) := \frac{1}{|\mathcal{I}|} \sum_{i \in \mathcal{I}} w_{a,i} \cdot \left(\gamma\!\left(a, i\right) - 1\right) \enspace \textrm{,}
  %JB: does the "-1" provide any benefits, or does it only shift scores?
  %TF: it is used to resemble the 0-indexed implementation. I'm also not happy with this, but using \lbrace 0, \dots, k - 1 \rbrace does also look ugly.
  %MI: Why is it so important to have this zero-based?
  \label{eq:rankingeq}
\end{equation}
with $w_{a,i} := 2$ if $\gamma\!\left(a, i\right) = k$ else $w_{a,i} := 1$.
To be applicable to $\hat{a}$, we extend \emph{score} by replacing $\gamma\!\left(a, i\right)$ by the model's predictions $f\!: \left(\mathcal{A} \cup \left\lbrace \hat{a} \right\rbrace\right) \times \mathcal{I} \rightarrow \left\lbrace 1, \dots, k \right\rbrace$.
%JB: as "$\operatorname{score}$" is not visibly different to the rest of the text, we should use some other way of highlighting
%TF: replaced it with emph{score} for now

@@

That discretized runtimes are sufficient for distinguishing solvers can also be shown by comparing the fraction of solver pairs whose runtimes are significantly different.
Within the SAT~Competition~2022 Anniversary Track, about \SI{89.95}{\%} of solver pairs are significantly different (according to a Wilcoxon-signed-rank test with $\alpha = 0.05$).
While only time-out labels (solver times out or not on a particular instance) are sufficient to significantly distinguish about \SI{87.04}{\%} of solver pairs, discrete-runtime labels are sufficient to distinguish about \SI{87.83}{\%} of solver pairs.
%\footnote{Based on hierarchical clustering with a log-single-link criterion for $k = 3$.}
This means that we do not lose much discriminatory power using discrete runtime labels.
%The number of significant comparisons between solvers \emph{alone}, though, does only offer limited insights as we might achieve significant differences in all pairs by assigning some randomly skewed values as labels.
%JB: thus, we should reorder this section to mention the argument "significant difference" (this paragraph) first and "deciding solver pairs" (previous paragraph, stronger argument) second (rather than giving the stronger argument first, making the second argument superfluous)
%TF: Removed the previous sentence
The predictiveness of the PAR-2 score, the labels' discriminatory power, as well as that it enables us to use classification, show the clear advantage of runtime discretization.
%JB: at the same time with what? with the other benefits?
%TF: removed redundant phrase

@@

\subsection{Instance Sampling}
\label{sec:main2}
Our instance-sampling strategies (also refer to Algorithm~\ref{algALBenchmark}; line 4) can be divided into two categories.
First, we present possible baseline approaches including existing work on instance-based selection strategies~\cite{Bossek021a,GentHJKMNN14,MatriconAFSH21}.
%JB: approaches under "Second" are also at least "inspired by existing work" and probably also pick "interesting" instances, so distinction between the two categories of approaches not entirely clear
%TF: made it clearer. First = instance-based methods, second = model-based methods
Second, we show promising machine-learning-based sampling strategies that make use of the model's label-prediction function $f$ and are inspired by existing work within the realms of active learning~\cite{settles2009active}.

@@

The model-based information-gain sampling strategy selects the instance with the highest expected entropy reduction.
%This implicitly assumes that the model's runtime-label probabilities for an instance are close to being correct.
%JB: not sure what "close to being correct" means excactly: does strategy assume correct predictions, but implicitly has a tolerance for wrong predictions, or does it explicitly account for errors somehow?
%TF: removed the previous sentence as it probably confuses readers more and is not essential for the explanation
To be more specific, we select the instance $i$ that maximizes
\begin{equation}
  \operatorname{IG}\!\left(i\right) := \operatorname{H}\!\left(i\right) - \sum_{n = 1}^{k} f'\!\left(\hat{a}, i\right)_{n} \operatorname{H}\!\left(i, n\right) \enspace \textrm{,}
\end{equation}
whereby $\operatorname{H}\!\left(i\right)$ denotes the entropy of the runtime labels $\gamma\!\left(a, i\right)$ for all $a \in \mathcal{A}$ and $\operatorname{H}\!\left(i, n\right)$ denotes the entropy of the runtime labels including the runtime label of $\hat{a}$ which is $n$.
The term $\operatorname{H}\!\left(i, n\right)$ is computed for every possible runtime label $n$.

@@

The Wilcoxon stopping criterion stops the iterative active-learning algorithm when we are on average certain enough that the predicted runtime labels are sufficiently different in comparison to the other solvers.
To be more specific, we stop when
\begin{equation}
  W_{\hat{a}} := \frac{1}{\left\lvert \mathcal{A} \right\rvert} \sum_{a \in \mathcal{A}} \operatorname{wilcoxon}_{\alpha}\!\left(\gamma\!\left(a, \mathcal{I}\right), f\!\left(\hat{a}, \mathcal{I}\right) \right)
\end{equation}
drops below a certain threshold.
%JB: if this is info gain regarding runtimes, doesn't Eq. (4) mean that we would prefer instances with low entropy of runtime distribution after considering the new solver, i.e., instances that are not very discriminatory?
%TF: I don't think so. We want to maximize the reduction in entropy of the runtime labels. Low entropy afterward -> we know that the solver behaves similarly to existing ones.
Thereby, \emph{wilcoxon} denotes the $p$-value of a Wilcoxon signed-rank test of the runtime-label distributions and $\gamma\!\left(a, \mathcal{I}\right)$ and $f\!\left(\hat{a}, \mathcal{I}\right)$ are the natural set extensions on $\gamma$ and $f$ respectively.
%JB: "set extension" might be rather technical and hard to understand; also, why "expected"? in the sense of "straightforward"?
%TF: I already saw this expression in other papers. I'm not sure how to say the same thing clearer in the same amount of space.
To improve the stability of this criterion, we use an exponential moving average
\begin{equation}
  W_{\exp}^{\left(m\right)} := \beta W_{\hat{a}} + \left(1 - \beta\right) W_{\exp}^{\left(m - 1\right)} \enspace \textrm{,}
\end{equation}
whereby $m$ denotes the count of iterations within Algorithm~\ref{algALBenchmark} and $W_{\exp}^{\left(0\right)} := 1$.

@@

    $O_{\operatorname{rt}} \leftarrow \left(\sum_{i \in \tilde{\mathcal{I}}} r\!\left(i, \hat{a}\right)\right) / \left(\sum_{i \in \mathcal{I}} r\!\left(i, \hat{a}\right)\right)$
    %JB: do we need to call a function here or can't we directly write how this is computed (define a variable starting at 0 to which the runtime fraction is added each iteration)?
    %TF: fixed

@@

\subsection{Hyper-parameters}
\label{sec:hyper}
Given Alg.~\ref{algALBenchmark}, there are several possible instantiations for the three phases, i.e., \emph{selection}, \emph{stopping}, and \emph{ranking}.
Also, there are different choices for the runtime-label prediction model.
Note that we are not considering all previously listed approaches since a grid search of all combinations would be infeasible.
Rather, we filter approaches based on preliminary experimental results (cf. Section~\ref{sec:evalprel}) and do the main end-to-end experiments only with a subset.
The end-to-end experiment configurations are given below.
%JB: still interesting to see which approaches were considered in preliminary experiments; could also list them here and indicate somehow which methods were only used preliminarily (e.g., by putting "(preliminary experiments only)" or some shorter label/sign behind them)
%TF: Added some selection methods that are mentioned.

>>>>

Our approach satisfies several desirable criteria for benchmarking:
It outputs a \emph{ranking} to compare the new solver against the existing set of solvers.
For this ranking, we do not even need to predict exact solver runtimes, which is trickier.
We optimize the \emph{runtime} that our strategy needs to arrive at its conclusion.
We use problem-instance and runtime \emph{features}.
Moreover, we select instances \emph{non-randomly} and \emph{incrementally}.
In particular, we take runtime information from already done experiments into account when choosing the next.
By doing so, we can control the properties of the benchmarking approach such as its required runtime.
Rather than solely outputting a binary classification, i.e, the new solver is worse than an existing solver, we provide a \emph{scoring} function that shows by which margin a solver is worse and how similar it is to existing solvers.
Our approach is \emph{scalable} in that it ranks a new solver $\hat{a}$ among any number of known solvers $\mathcal{A}$.
In particular, we only subsample the benchmark once instead of comparing pairwise against each other solver~\cite{MatriconAFSH21}.
Since a new solver idea is rarely best on the first try, it is desired to obtain a solver ranking fast.
In this way, a new solver idea can be discarded if it performs poorly across the board or may be further tweaked if it shows promising results at least in some cases.

We evaluate our approach with the SAT Competition~2022 Anniversary Track benchmark~\cite{sat2022}, consisting of 5355~instances and full runtimes of 28~solvers.
We perform cross-validation by treating each solver as new once and predicting these solvers' PAR-2 rank.
On average, our predictions reach about \SI{92}{\%} accuracy with only about \SI{10}{\%} of the runtime that would be needed to evaluate these solvers on the complete set of instances.

@@

Hoos et al. discuss which properties are most desirable when selecting SAT benchmark instances~\cite{HoosKSS13}.
The selection strategy presented is static, i.e., it does not depend on particular solvers to distinguish.
Selection criteria are instance variety to avoid over-fitting, adapted instance hardness (not too easy but also not too hard), and avoiding duplicate instances.
To filter too similar instances, they use a distance-based approach with the SATzilla features~\cite{XuHHL08,features}.
Their approach \emph{ranks} solvers, is \emph{feature-based}, and \emph{scalable} in the sense that the evaluation time scales linearly with more instances and solvers.
Also, it allows \emph{scoring} and thereby comparing solvers.
However, the approach does not optimize for benchmark \emph{runtime} and selects instances \emph{randomly}, apart from constraints on the instance hardness and feature distance.

@@

Matricon~et~al. present an incremental benchmark selection approach~\cite{MatriconAFSH21}.
Their \emph{per-set efficient algorithm selection problem} (PSEAS) is similar to our \emph{New-Solver Problem} (cf.~Definition~\ref{def:new-solver-problem}).
Given a pair of SAT solvers, they iteratively select a subset of instances until the desired confidence level is reached to decide which of the two solvers is better.
The selection of instances is dependent on the choice of the solvers to distinguish.
They calculate a scoring metric for all unselected instances, run the experiment with the highest score, and update the confidence.
Most of the proposed strategies in~\cite{MatriconAFSH21} are instance-based, i.e., no model of the solver runtimes is built.
Regarding our desired criteria of benchmark selection (Table~\ref{tab:requirements}), the aforementioned approach \emph{ranks} solvers, optimizes for \emph{runtime}, is \emph{feature-based}, and uses \emph{incremental non-random} sampling.
However, the approach only compares solvers binarily rather than providing a \emph{scoring}.
Thus, it is not clear how similar two given solvers are or on which instances they behave similarly.
Moreover, a major shortcoming is the lacking \emph{scalability} with the number of solvers.
Comparing only pairs of solvers, evaluating a new solver requires sampling several benchmarks.
In contrast, our approach allows comparing a new solver against a set of existing solvers with one benchmark.

@@

The posed \emph{New-Solver Problem} has stark similarities to the well-studied field of active learning AL within recommender systems, especially the \emph{new-user problem}~\cite{RubensESK15}.
On the one hand, we want to maximize the utility an instance provides to our model and, on the other hand, minimize the cost (CPU time) that is associated with its acquisition.
In contrast to traditional passive machine-learning methods with given instance labels, active learning allows for selecting instances for which to acquire labels.
AL algorithms can be categorized into \emph{synthesis-bas\-ed} \cite{0001AEMN22,GarzonMG22,2019gaal} and \emph{pool-bas\-ed} approaches \cite{GolbandiKL11,HarpaleY08,KorenBV09}.%SinhaED19,distribAL,TongK01,KapoorGUD07,KornerW06,MelvilleM04}.
While synthesis-based methods generate instances for labeling, pool-based methods rely on a fixed set of unlabeled instances from which to sample.

@@

Preliminary experiments, as well as research in portfolio solvers, have shown however that directly predicting runtimes without any transformation does not work well in practice~\cite{NgokoCT19,CollauttiMMO13}.
In our experiments, the mean squared error was within the same magnitude as the values to be predicted.
In contrast to that, discretizing runtimes and using those labels in the model's training and prediction yields good results (cf.~Section~\ref{sec:eval}).
%The following paragraphs motivate why it makes sense to discretize runtimes.%MI: no further motivation found, commented out
%JB: we should probably cut this a bit
%TF: I've commented out a few sentences but we probably need to cut more.

We use a function $\gamma\!: {\mathcal{A} \times \mathcal{I}} \rightarrow \left\lbrace 1, \dots, k \right\rbrace$ to transform runtime observations.
In our experiments, this function is a clustering algorithm.
Note that this transformation is done per instance to adapt to different instance difficulties. 
Also, we define ${\gamma\!\left(a, i\right) := k}$ iff $r\!\left(a, i\right) = \,\perp$ for all $a \in \mathcal{A}$ and $i \in \mathcal{I}$.
To determine a ranking based on those labels, we define
\begin{equation}
  \operatorname{score}\!\left(a\right) := \frac{1}{|\mathcal{I}|} \sum_{i \in \mathcal{I}} w_{a,i} \cdot \left(\gamma\!\left(a, i\right) - 1\right) \enspace \textrm{,}
  %JB: does the "-1" provide any benefits, or does it only shift scores?
  %TF: it is used to resemble the 0-indexed implementation. I'm also not happy with this, but using \lbrace 0, \dots, k - 1 \rbrace does also look ugly.
  %MI: Why is it so important to have this zero-based?
  \label{eq:rankingeq}
\end{equation}
with $w_{a,i} := 2$ if $\gamma\!\left(a, i\right) = k$ else $w_{a,i} := 1$.
To be applicable to $\hat{a}$, we extend \emph{score} by replacing $\gamma\!\left(a, i\right)$ by the model's predictions $f\!: \left(\mathcal{A} \cup \left\lbrace \hat{a} \right\rbrace\right) \times \mathcal{I} \rightarrow \left\lbrace 1, \dots, k \right\rbrace$.

@@

That discretized runtimes are sufficient for distinguishing solvers can also be shown by comparing the fraction of solver pairs whose runtimes are significantly different.
Within the SAT~Competition~2022 Anniversary Track, about \SI{89.95}{\%} of solver pairs are significantly different (according to a Wilcoxon-signed-rank test with $\alpha = 0.05$).
While only time-out labels (solver times out or not on a particular instance) are sufficient to significantly distinguish about \SI{87.04}{\%} of solver pairs, discrete-runtime labels are sufficient to distinguish about \SI{87.83}{\%} of solver pairs.
%\footnote{Based on hierarchical clustering with a log-single-link criterion for $k = 3$.}
This means that we do not lose much discriminatory power using discrete runtime labels.
The predictiveness of the PAR-2 score, the labels' discriminatory power, as well as that it enables us to use classification, show the clear advantage of runtime discretization.

@@

\subsection{Instance Sampling}
\label{sec:main2}
Our instance-sampling strategies (also refer to Algorithm~\ref{algALBenchmark}; line 4) can be divided into two categories.
First, we present possible baseline approaches including existing work on instance-based selection strategies~\cite{Bossek021a,GentHJKMNN14,MatriconAFSH21}.
Second, we show promising machine-learning-based sampling strategies that make use of the model's label-prediction function $f$ and are inspired by existing work within the realms of active learning~\cite{settles2009active}.

@@

The model-based information-gain sampling strategy selects the instance with the highest expected entropy reduction.
To be more specific, we select the instance $i$ that maximizes
\begin{equation}
  \operatorname{IG}\!\left(i\right) := \operatorname{H}\!\left(i\right) - \sum_{n = 1}^{k} f'\!\left(\hat{a}, i\right)_{n} \operatorname{H}\!\left(i, n\right) \enspace \textrm{,}
\end{equation}
whereby $\operatorname{H}\!\left(i\right)$ denotes the entropy of the runtime labels $\gamma\!\left(a, i\right)$ for all $a \in \mathcal{A}$ and $\operatorname{H}\!\left(i, n\right)$ denotes the entropy of the runtime labels including the runtime label of $\hat{a}$ which is $n$.
The term $\operatorname{H}\!\left(i, n\right)$ is computed for every possible runtime label $n$.

@@

The Wilcoxon stopping criterion stops the iterative active-learning algorithm when we are on average certain enough that the predicted runtime labels are sufficiently different in comparison to the other solvers.
To be more specific, we stop when
\begin{equation}
  W_{\hat{a}} := \frac{1}{\left\lvert \mathcal{A} \right\rvert} \sum_{a \in \mathcal{A}} \operatorname{wilcoxon}_{\alpha}\!\left(\gamma\!\left(a, \mathcal{I}\right), f\!\left(\hat{a}, \mathcal{I}\right) \right)
\end{equation}
drops below a certain threshold.
%JB: if this is info gain regarding runtimes, doesn't Eq. (4) mean that we would prefer instances with low entropy of runtime distribution after considering the new solver, i.e., instances that are not very discriminatory?
%TF: I don't think so. We want to maximize the reduction in entropy of the runtime labels. Low entropy afterward -> we know that the solver behaves similarly to existing ones.
Thereby, \emph{wilcoxon} denotes the $p$-value of a Wilcoxon signed-rank test of the runtime-label distributions and $\gamma\!\left(a, \mathcal{I}\right)$ and $f\!\left(\hat{a}, \mathcal{I}\right)$ are the natural set extensions on $\gamma$ and $f$ respectively.
To improve the stability of this criterion, we use an exponential moving average
\begin{equation}
  W_{\exp}^{\left(m\right)} := \beta W_{\hat{a}} + \left(1 - \beta\right) W_{\exp}^{\left(m - 1\right)} \enspace \textrm{,}
\end{equation}
whereby $m$ denotes the count of iterations within Algorithm~\ref{algALBenchmark} and $W_{\exp}^{\left(0\right)} := 1$.

@@

    $O_{\operatorname{rt}} \leftarrow \left(\sum_{i \in \tilde{\mathcal{I}}} r\!\left(i, \hat{a}\right)\right) / \left(\sum_{i \in \mathcal{I}} r\!\left(i, \hat{a}\right)\right)$

@@

\subsection{Hyper-parameters}
\label{sec:hyper}
Given Alg.~\ref{algALBenchmark}, there are several possible instantiations for the three phases, i.e., \emph{selection}, \emph{stopping}, and \emph{ranking}.
Also, there are different choices for the runtime-label prediction model.
Note that we are not considering all previously listed approaches since a grid search of all combinations would be infeasible.
Rather, we filter approaches based on preliminary experimental results (cf. Section~\ref{sec:evalprel}) and do the main end-to-end experiments only with a subset.
The end-to-end experiment configurations are given below.
